A Plasma Torus Around a Young Low-Mass Star

### 2025-01-18 paper/ms.tex
rename + new words

The two leading hypotheses to explain the CPVs are either that we are
in the shadow of transiting clumps of circumstellar material that
corotate with the star (CITE, CITE, CITE), or that these stars
represent an extreme in naturally-occurring distributions of starspots
or faculae for young M dwarfs \cite{Koen2021}.  Currently, the main
argument against a starspot-only explanation invokes the timescales
and amplitudes of the sharpest photometric features (CITE).  However,
there has yet to be any direct independent evidence for the presence of
circumstellar material in any of these objects.  Since transiting
circumstellar clumps would geometrically imply an occcurrence rate a
few to ten times the observed rate (CITE), this phenomenon has the
potential to affect 10-30\% of M dwarfs during their early lives.

The dearth of evidence for circumstellar material around CPVs is
surprising given that separate studies of young BAFGKM stars have, for
decades, reported that stellar coronae contain both hot ($10^6$ K) and
cool ($10^4$ K) plasma. In particular, time-series spectroscopy has
shown periodic high-velocity absorption and emission in Balmer lines
such as H$\alpha$, implying long-lived, corotating clumps of cool
plasma (CITE, CITE, CITE).  Such clumps are forced into corotation by
the magnetic field, and the geometry of where the plasma can
accumulate is thought to be dictated by the magnetic field's topology.
For instance, a tilted dipole field tends to yield an accumulation
surface of a warped torus \cite{Townsend2005}, whereas in the limit of
a single strong discrete field line, accumulation occurs along a fixed
point \cite{Waugh2022}.  However, none of these stars have shown any
photometric anomalies (CITE), leaving open the issue of whether these
two separate areas of study have any direct connection.  Nonetheless,
CPVs do respond to sudden magnetic field changes: there are many
documented cases of otherwise long-lived eclipse features disappearing
immediately following stellar flares (CITE, CITE).

In this study, we present the first observations of corotating clumps
of cool plasma around a CPV.  We identified TIC~141146667 in previous
work \cite{Bouma2024} by searching the TESS two-minute data for stars
showing periodic variability with at least three sharp dips.  We
selected it from the resulting fifty high-quality CPVs for follow-up
observations because it was the brightest source for which a full
cycle could be observed in a half-night.  We observed it for five
hours on UT 2024-02-17 using the High Resolution Echelle Spectrometer
(HIRES; \cite{vogt_hires_1994}) on the Keck I 10m telescope, roughly
contemporaneous with TESS, which observed the star from UT 2024-02-05
to UT 2024-02-26 with a duty cycle of XX\%.  In detail, TESS was
finishing a data downlink during the HIRES observations, and
photometric data collection resumed three cycle periods (12 hours)
after the spectra were acquired.  
Extended Data Figure~\ref{fig:fulllc} shows the detailed photometric
behavior of the star before and after the exact epoch of observation;
the star remained sufficiently stable to not affect any of the
interpretation that follows.

@@

@@

Figure~\ref{fig:spec} shows the available photometry and spectroscopy from
February 2024.
While the star's photometric shape changed from February 2022,
its behavior remained complex:
the average photometric signal shows a smooth brightening over 45\% of
the period, followed by a complex eclipse-like feature spanning 55\%
of the period.
This photometric eclipse shows two to three local minima, 
and one to two local maxima at times near the spectroscopic
observation.

The novelty is the spectroscopy.  Emission from material well beyond
the star's equatorial velocity ($v_{\rm eq}$=130\,\kms) is visible by
eye in the raw spectrogram.
The emission is complex...
%TODO: what are the features you need to quantify / quote?

slightly less than half the period,

While the two are not exactly simultaneous,

@@

These data rule out ``starspot-only'' and ``dust-only''
origin scenarios for CPVs, instead supporting either a purely stellar
origin for the phenomenon or extrinsic scenarios involving long-lived
disks or outgassing rocky bodies capable of supplying sufficient gas.

>>>>

The two leading hypotheses to explain the CPVs are either that
transiting clumps of circumstellar material corotate with the star
(CITE, CITE, CITE), or that these stars represent an extreme in
naturally-occurring distributions of starspots or faculae for young M
dwarfs \cite{Koen2021}.  Currently, the main argument against a
starspot-only explanation invokes the timescales and amplitudes of the
sharpest photometric features (CITE).  However, no independent
evidence has yet been presented for the presence of circumstellar
material in these objects.  Since transiting circumstellar
clumps would geometrically imply an occcurrence rate a few to ten
times the observed rate (CITE), the question of whether it is there
could potentially teach us about 10-30\% of M dwarfs during their
early lives.

@@

The dearth of evidence for circumstellar material around CPVs is
surprising given that separate studies of young BAFGKM stars have, for
decades, reported that stellar coronae contain both hot ($10^6$ K) and
cool ($10^4$ K) plasma. In particular, time-series spectroscopy has
shown periodic high-velocity absorption and emission in Balmer lines
such as H$\alpha$, caused by long-lived, corotating clumps of cool
plasma (CITE, CITE, CITE).  Such clumps are forced into corotation by
the magnetic field, and the exact geometry of where the plasma can
accumulate is dictated by the magnetic field's topology.
For instance, a tilted dipole field tends to yield an accumulation
surface of a warped torus \cite{Townsend2005}, whereas in the limit of
a single strong discrete field line, accumulation occurs along a fixed
point \cite{Waugh2022}.  However, none of these stars have shown any
photometric anomalies (CITE), leaving open the issue of whether these
two separate areas of study have any direct connection.  Nonetheless,
CPVs do respond to sudden magnetic field changes: there are many
documented cases of otherwise long-lived eclipse features disappearing
immediately following stellar flares (CITE, CITE).

In this study, we present the first observations of corotating clumps
of cool plasma around a CPV.  We identified TIC~141146667 in previous
work \cite{Bouma2024} by searching the TESS two-minute data for stars
showing periodic variability with at least three sharp dips per cycle.  We
selected it from the resulting fifty high-quality CPVs for
spectroscopic observations because it was the brightest source for
which a full cycle could be observed in a half-night.  We observed it
for five hours on UT 2024-02-17 using the High Resolution Echelle
Spectrometer (HIRES; \cite{vogt_hires_1994}) on the Keck I 10m
telescope, roughly contemporaneous with TESS, which observed the star
from UT 2024-02-05 to UT 2024-02-26 with a duty cycle of XX\%.  In
detail, TESS was finishing a data downlink during the HIRES
observations, and photometric data collection resumed three rotation cycles
(12 hours) after the spectra were acquired.  Extended Data
Figure~\ref{fig:fulllc} shows the detailed photometric behavior of the
star before and after the exact epoch of observation; the star
remained sufficiently stable to not affect any of the interpretation
that follows.

@@

% photometry
Figure~\ref{fig:spec} shows the data from February 2024.  As expected
based on other CPVs \cite{Bouma2024}, the photometric shape of
TIC~141146667 evolved over the years following the 2022 discovery
data, while nonetheless remaining complex.  In February 2024, the
average photometric signal showed a gradual brightening over 45\% of
the period, followed by a complex eclipse-like feature spanning 55\%
of the period.  This eclipse-like feature shows two to three
local photometric minima, and one to two local maxima.

% ** keck/hires: emission beyond v/veq>1 (2-3 paragraphs)
The spectroscopy shows emission well outside the star's
equatorial velocity ($v_{\rm eq}$=130\,\kms).  There are at least two
distinct emission components, each spaced half a cycle apart in phase.  The
first has clearer sinusoidal behaviour and is double-peaked, with
semi-amplitudes of $K_1$=2.1\,$v_{\rm eq}$ and 2.7\,$v_{\rm eq}$.  The
flux excesses from these two peaks are correlated with one another,
with amplitudes varying from 100\% of the continuum flux early in the
observation sequence to 30\% by its end.  The component 180$^\circ$
opposite in phase is only detected from $\phi$=0.2-1.0, and 
from $\phi$=0.2-0.5, this latter component appears connected to the
star in velocity space.  While its peak semi-amplitude of
$K_1$=3.9\,$v_{\rm eq}$ is achieved at both $\phi$=0.25 and 0.75, its
amplitude decreases from a 60\% excess beyond the continuum to a 10\%
excess.   The sinusoidal period in all cases where emission is seen is
consistent with the photometric \periodhr\ hour period.  As we shall
discuss below, the only plausible explanation for the behavior at
$|\Delta v / v_{\rm eq}|>1$ is that circumstellar clumps of hydrogen
are corotating with the star.  These clumps transit in front of the
star when passing from negative to positive velocity.

% * keck/hires: absorption within v/veq<1
% * keck/hires: emission within v/veq<1
Within the stellar H$\alpha$ line core, at $|\Delta v / v_{\rm
eq}|<1$, the behavior is more complex.  Generally, one would
expect this region to be generated by the net superposition of bright
and dark regions on the stellar surface, and then modulated by any
occulting material capable of absorbing or emitting in H$\alpha$.  In
Figure~\ref{fig:spec}e, the behavior from $\phi$=0.4-1.2, is most easily
interpreted: from $\phi$=0.4-0.9, a hot region first gradually crosses
the stellar line profile, following from $\phi$=0.7-1.2 by the transit
of a cool region.  The phases $\phi$$<$0.4 seem to be a mix of similar
events, though the time sampling is sufficiently coarse that the
interpretation is less clear.
A final exercise to quantify the behavior in the line core is
shown in Figure~\ref{fig:spec}d, where $f_{\rm H\alpha\ core}$ denotes
the summed flux in the at $|\Delta v / v_{\rm eq}|<1$.
Changes in the line core flux are usually correlated with the
broadband variability, except at $\phi$=0.5, during the transit of the
higher-velocity clump and the occultation of the lower-velocity clump.

@@

The sinusoidal emission features require clumps of partially-ionized
hydrogen to be corotating with the star.
The velocity semi-amplitude of the sinusoids directly gives the
distance of these clumps from the stellar surface: 2.1-2.7\,$R_\star$
for one clump, and 3.9\,$R_\star$ for the other.
Their motion, rather than being Keplerian, can only be explained by
plasma being dragged along with the rotating stellar magnetic field.

The new data rule out two plausible origin scenarios for CPVs.
The ``starspot-only'' scenario \cite{Koen2021} has no means of
explaining this type of spectroscopic emission.
Similarly, a ``dust-only'' scenario in which the circumstellar
material would have a pure dust composition is also ruled out:
to explain the H$\alpha$ emission, the circumstellar clumps need to be
made of either a pure plasma, or a dusty plasma.
A priori, there are three plausible sources of such material:
the star, an old and undetected disk, or outgassing rocky bodies.

% * mass estimate
The density and mass of the material...

% TODO TODO: the rest!
* role of B: comparison to, and need for, RMHD models
  * massive B star / jupiter cnxn

* more wavelengths:
    Radio detection.   Relevance for B measurement.
    UV?  xray?
    need for jwst

* microphysics: the (probable) need for dust
    W1,W2,W3 no dust evidence.
    W4 detection questionable based on WISE images.
